---
title: "Networks - day2"
author: "CB"
date: "2024-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Visualization of networks

## The MDS algorithm

```{r}
A = cbind(c(0,1,1,1,0),
          c(1,0,0,0,1),
          c(1,0,0,0,1),
          c(1,0,0,0,1),
          c(0,1,1,1,0))
image(A)
```
> Exercise: define a function `mds.loss` that encodes the loss function of MDS to minimize and then call an optimization algorithm to run MDS on the above data



```{r}
library(igraph); library(sna)

mds.loss <-function(z,A){
  # A is the adjacency matrix
  # Z are the 2-dimensional latent position
  # (provided as a vector)
  g = graph_from_adjacency_matrix(as.matrix(A))
  D.in = shortest.paths(g)
  Z = matrix(z,ncol = 2)
  D.out = as.matrix(dist(Z)) # The Euclidean distance in R^2 
  loss = sum((D.in - D.out)^2)
}

# Optimization of the MDS loss within a MDS function
my.MDS <- function(A){
  n = nrow(A)
  z.init = runif(2*n,-1,1)
  out = optim(z.init,mds.loss,A=A,method="SANN")
  list(Z.final = matrix(out$par,ncol = 2),
         loss = out$value,
         Z.init = matrix(z.init,ncol=2))
}

# Running the MDS algorithm
res = my.MDS(A)

# Visualization of the latent representation
par(mfrow=c(1,2))
gplot(A,coord=res$Z.init,edge.col="gray")
title(main = "Initialization of MDS")
gplot(A,coord=res$Z.final,edge.col="gray")
title(main=paste("MDS solution (loss=",round(res$loss,3),")",sep=''))
```
For simpler networks, the loss can be very low:

```{r}
A = cbind(c(0,1,1,1),
          c(1,0,1,0),
          c(1,1,0,0),
          c(1,0,0,0))

# Running the MDS algorithm
res = my.MDS(A)

# Visualization of the latent representation
par(mfrow=c(1,2))
gplot(A,coord=res$Z.init,edge.col="gray")
title(main = "Initialization of MDS")
gplot(A,coord=res$Z.final,edge.col="gray")
title(main=paste("MDS solution (loss=",round(res$loss,3),")",sep=''))
```
Let's now apply this method on the Sampson Monk data:

```{r}
#install.packages("VBLPCM")
library(VBLPCM)
data(sampson)
library(network)
A = as.matrix.network.adjacency(samplike)

# Running the MDS algorithm
res = my.MDS(A)

# Visualization of the latent representation
par(mfrow=c(1,2))
gplot(A,coord=res$Z.init,edge.col="gray")
title(main = "Initialization of MDS")
gplot(A,coord=res$Z.final,edge.col="gray")
title(main=paste("MDS solution (loss=",round(res$loss,3),")",sep=''))
```

## The latent space model

> Write a function `my.LSM` which implements the LSM algorithm.

```{r}
LSM.loss <- function(par,A){
  alpha = par[1]
  Z = matrix(par[-1],ncol=2)
  D.out = as.matrix(dist(Z))
  log.lik =  sum(A * (alpha - D.out) - log(1+exp(alpha - D.out)))
  return(-log.lik)
}

my.LSM <- function(A){
  n = nrow(A)
  alpha.init = 0
  z.init = runif(2*n,-1,1)
  par.init = c(alpha.init,z.init) 
  out = optim(par.init,LSM.loss,A=A,method="SANN")
  list(alpha = out$par[1],
      Z.final = matrix(out$par[-1],ncol = 2),
      loss = out$value,
      Z.init = matrix(z.init,ncol=2))
}

A = cbind(c(0,1,1,1),
          c(1,0,1,0),
          c(1,1,0,0),
          c(1,0,0,0))

# Running the LSM algorithm
res = my.LSM(A)

# Visualization of the latent representation
par(mfrow=c(1,2))
gplot(A,coord=res$Z.init,edge.col="gray")
title(main = "Initialization of MDS")
gplot(A,coord=res$Z.final,edge.col="gray")
title(main=paste("LSM solution (loss=",round(res$loss,3),")",sep=''))
```

Let's now apply this function to the Sampson data:

```{r}
library(VBLPCM); library(network)
data(sampson)
A = as.matrix.network.adjacency(samplike)

# Running the MDS and LSM algorithm
res.MDS = my.MDS(A)
res.LSM = my.LSM(A)

# Visualization of the latent representation
par(mfrow=c(1,2))
gplot(A,coord=res.MDS$Z.final,edge.col="gray")
title(main = "MDS solution")
gplot(A,coord=res.LSM$Z.final,edge.col="gray")
title(main = "LSM solution")
```

